Using functor categories to generate intermediate code with Agda

### 2025-05-11 draft.tex
Evaluation almost done

\documentclass[12pt,twoside,a4paper]{report}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathabx}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{enumerate}
\usepackage[vmargin=2.5cm, hmargin=2.5cm]{geometry}
\usepackage{hyperref}
\usepackage{fontspec}
\usepackage{unicode-math}
\usepackage{stackengine}
\usepackage{minitoc}
\usepackage{amsthm}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{xeCJK}
\usepackage{dirtree}
\usepackage{float}
\usepackage{pdfpages}
\usepackage{atbegshi}

@@

\newcommand{\bN}{\mathbb{N}}

@@

\newcommand{\bracket}[1]{\left\{ #1 \right\}}
\newcommand{\ang}[1]{\left\langle #1 \right\rangle}

@@

@@

    \begin{tabular}{ll}
        \textbf{Candidate Number:} & 2330G \\
        \textbf{Title of Project:} & Using functor categories to generate intermediate code with Agda \\
        \textbf{Examination} & Computer Science Tripos - Part II - 2025 \\
        \textbf{Word-Count:} & [wordcount] \\
        \textbf{Code line count:} & [linecount] \\
        \textbf{Project Originator:} & Yulong Huang \\
        \textbf{Project Supervisor:} & Yulong Huang and Yufeng Li \\
    \end{tabular}

@@

    \vspace{2em}
    \textbf{\Large Special Difficulties}
    \vspace{1em}
\end{titlepage}

@@

\mainmatter
\chapter{Introduction}
    \minitoc
    Programming languages act as a bridge between human thought and machine execution. They exist in two complementary realms: the human realm, where intent is expressed through abstractions like variables and functions, and the machine realm, where low-level instructions are executed on hardware.

@@

    This project explores how the two process can be deeply connected by demonstrating how denotational semantics can directly generate a compiler. Simple Typed Lambda Calculus (STLC) is a well-studied programming language that serves as a foundation for many modern languages. It has denotational semantics in cartesian closed categories, shown by Lambek~\cite{lambek}. As a ccc, presheaf categories over store locations can be used to model STLC with stores, as shown by by Reynolds~\cite{essence} and Oles~\cite{Oles_1}~\cite{Oles_2}. Instead, Reynolds~\cite{Reynolds} presented a denotational semantics of STLC with stores in the form of a presheaf category over compiler states. By interpresting the source language into the presheaf category over stack descriptors, where objects of the category represent instruction sequences parameterised by stack layouts, the semantic model directly yields a compiler.

    I implement this compiler for STLC with stores in a dependently typed programming language, Agda~\cite{Agda}. The compiler is a functor (structure-preserving map) from the source language's semantic domain to the target language's instruction set. The implementation is verified with Agda's type system, which ensures that the generated code is well-typed and adheres to the semantics of both the source and target languages.

@@

        % \motivation{Trend of verified compilers}
        % The rise of verified compilers including CompCert~\cite{CompCert}, CakeML~\cite{CakeML} reflects a broader trend toward trustworthy systems, where correctness proofs replace testing for critical guarantees. Like Lean 4~\cite{Lean4}, we leverage dependent types to internalise the verification of correctness of terms.

@@

        Agda also provides an interactive environment for writing and verifying programs, which will be further discussed in \secref{subsec: holes}.

    \section{Contributions}
        Addressed the three motivations presented in \secref{sec: motivation} and contributed to the following:
        \begin{quote}
            \savedmotivationI
            I implemented the compiler from the source language to the target language in Agda.

            \savedmotivationII
            I formalised the terms in the source language and target language in Agda, and proved that the compiler is a functor from the source language to the target language.

            % \savedmotivationIII
            % This implementation follows the trend of verified compilers, and provides a practical example of how to use functor categories to generate intermediate code.
        \end{quote}

@@

        Notation in category theory is similar to that in set theory and type theory. For example, $A, B, \dots$ are used to denote objects, sets, or types, and arrows are used to denote morphisms or functions (e.g.\ $A \to B$). There is a deeper connection between category theory and type theory, which will 
        be discussed in \secref{subsec: curry-howard-lambek}.

@@

        \subsection{Category}
        \begin{definition}[Category]
            A \emph{category} $\mathcal{C}$ is specified by the following:
            \begin{itemize}
                \item 
                    a collection of objects $\textbf{obj}(\mathcal{C})$, whose elements are called $\mathcal{C}$-objects;
                \item 
                    for each $X, Y \in \textbf{obj}(\mathcal{C})$, a collection of morphisms $\mathcal{C}{(X,Y)}$, whose elements are called $\mathcal{C}$-morphisms from $X$ to $Y$ (e.g.\ $f : X \to Y$);
                \item 
                    for each $X \in \textbf{obj}(\mathcal{C})$, an element $\textbf{id}_X \in \mathcal{C}{(X,X)}$ called the identity morphism on $X$;
                \item 
                    for each $X, Y, Z \in \textbf{obj}(\mathcal{C})$, a function 
                    \[\begin{aligned}
                        \mathcal{C}{(X,Y)} \times \mathcal{C}{(Y,Z)} &\to \mathcal{C}{(X,Z)} \\
                        (f,g) &\mapsto g \circ f
                    \end{aligned}\]
                    called the composition of morphisms;
            \end{itemize}
            satisfying the following properties:
            \begin{itemize}
                \item 
                    \textbf{(Unit Laws)}
                    for all $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, we have:
                    \begin{equation} \label{law: unit}
                        \textbf{id}_Y \circ f = f = f \circ \textbf{id}_X
                    \end{equation}
                \item
                    \textbf{(Associativity Law)}
                    for all $X, Y, Z, W \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $g \in \mathcal{C}{(Y,Z)}$, $h \in \mathcal{C}{(Z,W)}$, we have:
                    \begin{equation} \label{law: associativity}
                        h \circ (g \circ f) = (h \circ g) \circ f
                    \end{equation}
            \end{itemize}
        \end{definition}

        \begin{example}[$\mathcal{Set}$]
            An example of category is the category of sets, denoted as $\mathcal{Set}$, specifed by:
            \begin{itemize}
                \item 
                    the objects of $\mathcal{Set}$ are a fixed universe of sets;
                \item
                    for each $X, Y \in \textbf{obj}(\mathcal{Set})$, the morphisms from $X \to Y$ in $\mathcal{Set}$ are the functions $X \to Y$;
                \item
                    the identity morphism on $X$ in $\mathcal{Set}$ is the identity function on $X$;
                \item
                    the composition of morphisms in $\mathcal{Set}$ is defined as the composition of functions.
            \end{itemize}
            Associativity law and unit laws are satisfied in $\mathcal{Set}$, since the composition of functions is associative and the identity function compose with any function is the function itself.
        \end{example}

        \begin{example}[$\mathcal{Poset}$]
            Another example of category is the category of posets, denoted as $\mathcal{Poset}$, specified by:
            \begin{itemize}
                \item 
                    the objects of $\mathcal{Poset}$ are a fixed universe of posets, which is a set $P$ with a binary relation $\subseteq$ that is 
                    \begin{itemize}
                        \item reflexive: $\forall x \in P, x \subseteq x$;
                        \item transitive: $\forall x, y, z \in P, x \subseteq y \land y \subseteq z \implies x \subseteq z$;
                        \item antisymmetric: $\forall x, y \in P, x \subseteq y \land y \subseteq x \implies x = y$.
                    \end{itemize}
                \item
                    for each $X, Y \in \textbf{obj}(\mathcal{Poset})$, the morphisms from $X$ to $Y$ in $\mathcal{Poset}$ are the monotone (order-preserving) functions from $X$ to $Y$;
                \item
                    the identity morphism on $X$ in $\mathcal{Poset}$ is the identity function on $X$;
                \item
                    the composition of morphisms in $\mathcal{Poset}$ is defined as the composition of functions.
            \end{itemize}
            Associativity law and unit laws are satisfied in $\mathcal{Poset}$, since the set of monotone functions contains identity functions and is closed under composition.

            $\mathcal{Poset}$ is used later for stack descriptors in the denotational semantics of the source language.
        \end{example}

@@

            \begin{definition}[Opposite category]
                Given a category $\mathcal{C}$, its \emph{opposite category} $\mathcal{C}^{\textbf{op}}$ is specified by:
                \begin{itemize}
                    \item 
                        the objects of $\mathcal{C}^{\textbf{op}}$ are the same as those of $\mathcal{C}$;
                    \item 
                        for each $X, Y \in \textbf{obj}(\mathcal{C})$, the morphisms from $X$ to $Y$ in $\mathcal{C}^{\textbf{op}}$ are the morphisms from $Y$ to $X$ in $\mathcal{C}$;
                    \item 
                        the identity morphism on $X$ in $\mathcal{C}^{\textbf{op}}$ is the identity morphism on $X$ in $\mathcal{C}$;
                    \item 
                        the composition of morphisms in $\mathcal{C}^{\textbf{op}}$ is defined as the composition of morphisms in $\mathcal{C}$.
                \end{itemize}
            \end{definition}

@@

            As examples of commutative diagrams, Figure \ref{fig: associativity_law} and Figure \ref{fig: unit_laws} are commutative diagrams for the unit laws and associativity law respectively.

@@

        \subsection{Cartesian closed category}
            \subsubsection{Terminal object}
            \begin{definition}[Terminal object]
                Given a category $\mathcal{C}$, an object $T \in \textbf{obj}(\mathcal{C})$ is called a \emph{terminal object} if for all $X \in \textbf{obj}(\mathcal{C})$, there exists a unique \textbf{C}-morphism $f : X \to T$.
            \end{definition}
            Terminal objects are unique up to isomorphism. In other words, we have
            \begin{itemize}
                \item 
                    if $T$ and $T'$ are both terminal objects in $\mathcal{C}$, then there exists a unique isomorphism $f : T \to T'$.
                \item
                    if $T$ is a terminal object in $\mathcal{C}$ and $T \cong T'$, then $T'$ is also a terminal object in $\mathcal{C}$.
            \end{itemize}
            \begin{example}[Terminal object in $\mathcal{Set}$]
                $\mathcal{Set}$ has a terminal object $\{\ast\}$, which is an arbitrary singleton set containing a single element $\ast$.

@@

            \subsubsection{Binary product}
            \begin{definition}[Binary product]
                Given a category $\mathcal{C}$, the \emph{binary product} of two objects $X$ and $Y$ in $\mathcal{C}$ is specified by
                \begin{itemize}
                    \item a $\mathcal{C}$-object $X \times Y$;
                    \item two $\mathcal{C}$-morphisms $\pi_1 : X \times Y \to X$ and $\pi_2 : X \times Y \to Y$ called the \emph{projections} of $X \times Y$;
                \end{itemize}
                such that for all $Z \in \textbf{obj}(\mathcal{C})$ and morphisms $f : Z \to X$ and $g : Z \to Y$, there exists a unique morphism $u : Z \to X \times Y$ such that the following diagram commutes in $\mathcal{C}$:
                \begin{equation}
                    \vcenter{\hbox{\includegraphics[width=0.4\textwidth]{binary_product.pdf}}}
                    \label{cd: binary_product}
                \end{equation}
                The unique morphism $u$ is written as 
                \[ \bracket{ f, g } : Z \to X \times Y \]
                where $f = \pi_1 \circ u$ and $g = \pi_2 \circ u$.
            \end{definition}
            It can be shown that the binary product is unique up to (unique) isomorphism.

@@

        \subsection{Functor}
        \begin{definition}[Functor]
            Given two categories $\mathcal{C}$ and $\mathcal{D}$, a \emph{functor} $F: \mathcal{C} \to \mathcal{D}$ is specified by:
            \begin{itemize}
                \item 
                    a function 
                    \[\begin{aligned}
                        \textbf{obj}(\mathcal{C}) &\to \textbf{obj}(\mathcal{D}) \\
                        X &\mapsto F(X)
                    \end{aligned}\]

                \item 
                    for each $X, Y \in \textbf{obj}(\mathcal{C})$, a function 
                    \[\begin{aligned}
                        \mathcal{C}{(X,Y)} &\to \mathcal{D}{(F(X),F(Y))} \\
                        f &\mapsto F(f)
                    \end{aligned}\]
            \end{itemize}
            satisfying the following properties:
            \begin{itemize}
                \item 
                    for all $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, we have: $ F(\textbf{id}_X) = \textbf{id}_{F(X)}$
                \item
                    for all $X, Y, Z \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $g \in \mathcal{C}{(Y,Z)}$, we have: $F(g \circ f) = F(g) \circ F(f)$
            \end{itemize}
        \end{definition}

@@

        \subsection{Functor category}
        \begin{definition}[Functor category]
            Given two categories $\mathcal{C}$ and $\mathcal{D}$, the \emph{functor category} $\mathcal{D}^{\mathcal{C}}$ is the category satisfying the following:
            \begin{itemize}
                \item 
                    the objects of $\mathcal{D}^{\mathcal{C}}$ are all functors $\mathcal{C} \to \mathcal{D}$;
                \item 
                    given two functors $F,G: \mathcal{C} \to \mathcal{D}$, the morphisms from $F$ to $G$ in $\mathcal{D}^{\mathcal{C}}$ are all natural transformations $\theta : F \to G$;
                \item
                    composition and identity morphisms in $\mathcal{D}^{\mathcal{C}}$ are defined as follows:
                    \begin{itemize}
                        \item 
                            the identity morphism $\textbf{id}_F$ on $F$ is defined as $\theta_X = \textbf{id}_{F(X)}$ for all $X \in \textbf{obj}(\mathcal{C})$;
                        \item
                            the composition of two natural transformations $\theta : F \to G$ and $\phi : G \to H$ is defined as $(\phi \circ \theta)_X = \phi_X \circ \theta_X$ for all $X \in \textbf{obj}(\mathcal{C})$.
                    \end{itemize}
            \end{itemize}
        \end{definition}

        \subsection{Presheaf category}
        \begin{definition}[Presheaf]
            Given a category $\mathcal{C}$, a \emph{presheaf} on $\mathcal{C}$ is a functor $F: \mathcal{C}^{\textbf{op}} \to \textbf{Set}$.
            A presheaf is a contravariant functor, which means that it reverses the direction of morphisms.
            In other words, a presheaf is a functor that takes objects in $\mathcal{C}$ and assigns them sets, and takes morphisms in $\mathcal{C}$ and assigns them functions between the corresponding sets.
            The presheaf $F$ is defined as follows:
            \begin{itemize}
                \item 
                    for each $X \in \textbf{obj}(\mathcal{C})$, $F(X)$ is a set;
                \item
                    for each $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $F(f)$ is a function $F(Y) \to F(X)$.
            \end{itemize}
        \end{definition}

        \begin{definition}[Presheaf category]
            Given a category $\mathcal{C}$, the \emph{presheaf category} $\hat{\mathcal{C}}$ is the is the functor category $\mathcal{Set}^{\mathcal{C}^{\textbf{op}}}$, which explicitly contains the following:
            \begin{itemize}
                \item 
                    the objects of $\hat{\mathcal{C}}$ are all presheaves on $\mathcal{C}$;
                \item
                    given two presheaves $F,G: \mathcal{C}^{\textbf{op}} \to \mathcal{Set}$, the morphisms from $F$ to $G$ in $\hat{\mathcal{C}}$ are all natural transformations $\theta : F \to G$.
            \end{itemize}
        \end{definition}

        \subsection{Yoneda lemma}
        \begin{definition}[Yoneda functor]
            Given a category $\mathcal{C}$, the \emph{Yoneda functor} $\yo: \mathcal{C} \to \hat{\mathcal{C}}$ is defined as follows:
            \begin{itemize}
                \item 
                    for each $X \in \textbf{obj}(\mathcal{C})$, $\yo(X)$ is the functor $\mathcal{C}^{\textbf{op}} \to \mathcal{Set}$ defined as:
                    \begin{equation} \label{eq: yoneda_element}
                        \yo(X)(Y) = \mathcal{C}{(Y,X)}
                    \end{equation}
                    for all $Y \in \textbf{obj}(\mathcal{C})$;
                \item
                    for each $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $\yo(f)$ is the morphism $\yo(X) \to \yo(Y)$ defined a natural transformation whose component at any given $Z \in \mathcal{C}^{\textbf{op}}$ is given by:
                    \begin{equation} \label{eq: yoneda_morphism}
                        \begin{aligned}
                            (\yo(f))_Z : \mathcal{C}{(Z,Y)} &\to \mathcal{C}{(Z,X)} \\
                            g &\mapsto g \circ f
                        \end{aligned}
                    \end{equation}
                    for all $Z \in \textbf{obj}(\mathcal{C})$.
                \end{itemize}

@@

        Here we define a datatype $\mb{\mathsf{Vec}}$ for vectors. The type of vectors is dependent on the length of the vector. With dependent types, we can encode properties directly into types and ensure that they are satisfied at compile time.

>>>>

\documentclass[12pt,twoside,a4paper]{report}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathabx}
\usepackage{stmaryrd}
\usepackage{upgreek}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{enumerate}
\usepackage[vmargin=2.5cm, hmargin=2.5cm]{geometry}
\usepackage{hyperref}
\usepackage{fontspec}
\usepackage{unicode-math}
\usepackage{stackengine}
\usepackage{minitoc}
\usepackage{amsthm}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{xeCJK}
\usepackage{dirtree}
\usepackage{float}
\usepackage{pdfpages}
\usepackage{atbegshi}
\usepackage{multirow, makecell}
\usepackage{enumitem}
\usepackage{hhline}

@@

\newcommand{\bN}{\mathbb{N}}
\newcommand{\bZ}{\mathbb{Z}}

@@

\newcommand{\bracket}[1]{\left\{ #1 \right\}}
\newcommand{\ang}[1]{\left\langle #1 \right\rangle}
\newcommand{\intp}[1]{\left\llbracket #1 \right\rrbracket}

@@

\renewcommand{\arraystretch}{1.5}

@@

    \begin{tabular}{ll}
        \textbf{Candidate Number:} & 2330G \\
        \textbf{Title of Project:} & Using functor categories to generate intermediate code with Agda \\
        \textbf{Examination} & Computer Science Tripos - Part II - 2025 \\
        \textbf{Word-count:} & [wordcount] \footnotemark \\
        \textbf{Code line count:} & [linecount] \footnotemark \\
        \textbf{Project Originator:} & Yulong Huang \\
        \textbf{Project Supervisor:} & Yulong Huang and Yufeng Li \\
    \end{tabular}

@@

    \vspace{2em}
    \textbf{\Large Special Difficulties}
    \vspace{1em}
    \begingroup
        \footnotetext[1]{This word-count was computed by \texttt{texcount -1 -sum -merge -q dissertation.tex}.}
        \footnotetext[2]{This code line count was computed by \texttt{find . -name "*.agda" | xargs wc -l}.}
    \endgroup
\end{titlepage}

@@

\mainmatter
\chapter{Introduction}
    \minitoc
    Programming languages act as a bridge between human thoughts and machine execution. They exist in two complementary realms: the human realm, where intent is expressed through abstractions like variables and functions, and the machine realm, where low-level instructions are executed on hardware.

@@

    This project explores how the two process can be deeply connected by demonstrating how denotational semantics can directly generate a compiler. Simply typed lambda calculus (STLC)~\cite{stlc} is a well-studied programming language that serves as a foundation for many modern languages. It has denotational semantics in cartesian closed categories, shown by Lambek~\cite{lambek}. As a ccc, presheaf categories over store locations can be used to model STLC with stores, as shown by by Reynolds~\cite{essence} and Oles~\cite{Oles_1}~\cite{Oles_2}. Instead, Reynolds~\cite{Reynolds} presented a denotational semantics of STLC with stores in the form of a presheaf category over compiler states. By interpresting the source language into the presheaf category over stack descriptors, where objects of the category represent instruction sequences parameterised by stack layouts, the semantic model directly yields a compiler.

    In this dissertation, I implement this compiler for STLC with stores in a dependently typed programming language, Agda~\cite{Agda}. The compiler is a functor (structure-preserving map) from the source language's semantic domain to the target language's instruction set. The implementation is verified with Agda's type system, which ensures that the generated code is well-typed and adheres to the semantics of both the source and target languages.

@@

@@

        Agda also provides an interactive environment for writing and verifying programs, which will be further discussed in~\secref{subsec: holes}.

    \section{Contributions}
        Addressed the two motivations presented in~\secref{sec: motivation} and contributed to the following:
        \begin{quote}
            \savedmotivationI
            I implemented the compiler from the source language to the target language in Agda.

            \savedmotivationII
            I formalised the terms in the source language and target language in Agda. I also refined some type definitions in the denotational semantics of the source language.

        \end{quote}

@@

        Notation in category theory is similar to that in set theory and type theory. For example, $A, B, \dots$ are used to denote objects, sets, or types, and arrows are used to denote morphisms or functions (e.g.\ $A \to B$). There is a deeper connection between category theory and type theory, which will be discussed in~\secref{subsec: curry-howard-lambek}.

@@

        \subsection{Category}
        \begin{definition}[Category]
            A \emph{category} $\mathcal{C}$ is specified by
            \begin{itemize}
                \item 
                    a collection of objects $\textbf{obj}(\mathcal{C})$, whose elements are called $\mathcal{C}$-objects;
                \item 
                    for each $X, Y \in \textbf{obj}(\mathcal{C})$, a collection of morphisms $\mathcal{C}{(X,Y)}$, whose elements are called $\mathcal{C}$-morphisms from $X$ to $Y$ (e.g.\ $f : X \to Y$);
                \item 
                    for each $X \in \textbf{obj}(\mathcal{C})$, an element $\textbf{id}_X \in \mathcal{C}{(X,X)}$ called the identity morphism on $X$;
                \item 
                    for each $X, Y, Z \in \textbf{obj}(\mathcal{C})$, a function 
                    \[\begin{aligned}
                        \mathcal{C}{(X,Y)} \times \mathcal{C}{(Y,Z)} &\to \mathcal{C}{(X,Z)} \\
                        (f,g) &\mapsto g \circ f
                    \end{aligned}\]
                    called the composition of morphisms;
            \end{itemize}
            satisfying the following properties:
            \begin{itemize}
                \item 
                    \textbf{(Unit Laws)}
                    For all $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, we have
                    \begin{equation} \label{law: unit}
                        \textbf{id}_Y \circ f = f = f \circ \textbf{id}_X
                    \end{equation}
                \item
                    \textbf{(Associativity Law)}
                    For all $X, Y, Z, W \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $g \in \mathcal{C}{(Y,Z)}$, $h \in \mathcal{C}{(Z,W)}$, we have
                    \begin{equation} \label{law: associativity}
                        h \circ (g \circ f) = (h \circ g) \circ f
                    \end{equation}
            \end{itemize}
        \end{definition}

        \begin{example}[$\mathcal{Set}$]
            An example of category is the category of sets, denoted as $\mathcal{Set}$, specifed by the following:
            \begin{itemize}
                \item 
                    The objects of $\mathcal{Set}$ are a fixed universe of sets;
                \item
                    For each $X, Y \in \textbf{obj}(\mathcal{Set})$, the morphisms from $X \to Y$ in $\mathcal{Set}$ are the functions $X \to Y$;
                \item
                    The identity morphism on $X$ in $\mathcal{Set}$ is the identity function on $X$;
                \item
                    The composition of morphisms in $\mathcal{Set}$ is defined as the composition of functions.
            \end{itemize}
            Associativity law and unit laws are satisfied in $\mathcal{Set}$, since the composition of functions is associative and the identity function compose with any function is the function itself.
        \end{example}

        \begin{definition}[Preorder]
            A \emph{preorder} $\underline{P} = (P, \sqsubseteq)$ is a set $P$ with a binary relation $\sqsubseteq$ that is
            \begin{itemize}
                \item reflexive: $\forall x \in P, x \sqsubseteq x$;
                \item transitive: $\forall x, y, z \in P, x \sqsubseteq y \land y \sqsubseteq z \implies x \sqsubseteq z$.
            \end{itemize}
        \end{definition}

        \begin{example}[Category determined by preorder] \label{ex: category_preorder}
            Another example of category is a category $\mathcal{C}_{\underline{P}}$ determined by any preorder $\underline{P}$, which is specified by the following:
            \begin{itemize}
                \item 
                    The objects of $\mathcal{C}_{\underline{P}}$ are the elements of $P$;
                \item
                    $\mathcal{C}_{\underline{P}}(x,y) = \begin{cases}
                        \{(x,y)\} & \text{if $x \sqsubseteq y$} \\
                        \emptyset & \text{otherwise}
                    \end{cases}$
                \item
                    The identity morphism on $X$ in $\mathcal{C}_{\underline{P}}$ is the identity function on $X$;
                \item
                    The composition of morphisms in $\mathcal{C}_{\underline{P}}$ is defined as the composition of functions.
            \end{itemize}
            Associativity law and unit laws are satisfied in $\mathcal{C}_{\underline{P}}$.

            $\mathcal{C}_{\underline{P}}$ is used later for stack descriptors in the denotational semantics of the source language.
        \end{example}

@@

            \begin{definition}[Opposite category]
                Given a category $\mathcal{C}$, its \emph{opposite category} $\mathcal{C}^{\textbf{op}}$ is specified by the following:
                \begin{itemize}
                    \item 
                        The objects of $\mathcal{C}^{\textbf{op}}$ are the same as those of $\mathcal{C}$;
                    \item 
                        For each $X, Y \in \textbf{obj}(\mathcal{C})$, the morphisms from $X$ to $Y$ in $\mathcal{C}^{\textbf{op}}$ are the morphisms from $Y$ to $X$ in $\mathcal{C}$;
                    \item 
                        The identity morphism on $X$ in $\mathcal{C}^{\textbf{op}}$ is the identity morphism on $X$ in $\mathcal{C}$;
                    \item 
                        The composition of morphisms in $\mathcal{C}^{\textbf{op}}$ is defined as the composition of morphisms in $\mathcal{C}$.
                \end{itemize}
            \end{definition}

@@

            As examples of commutative diagrams, Figure~\ref{fig: associativity_law} and Figure~\ref{fig: unit_laws} are commutative diagrams for the unit laws and associativity law respectively.

@@

        \subsection{Cartesian closed category}
            \subsubsection{Terminal object}
            \begin{definition}[Terminal object]
                Given a category $\mathcal{C}$, an object $T \in \textbf{obj}(\mathcal{C})$ is called a \emph{terminal object} if for all $X \in \textbf{obj}(\mathcal{C})$, there exists a unique \textbf{C}-morphism $f : X \to T$.
            \end{definition}
            Terminal objects are unique up to isomorphism. In other words, we have the following properties:
            \begin{itemize}
                \item 
                    If $T$ and $T'$ are both terminal objects in $\mathcal{C}$, then there exists a unique isomorphism $f : T \to T'$.
                \item
                    If $T$ is a terminal object in $\mathcal{C}$ and $T \cong T'$, then $T'$ is also a terminal object in $\mathcal{C}$.
            \end{itemize}
            \begin{example}[Terminal object in $\mathcal{Set}$]
                $\mathcal{Set}$ has a terminal object $\{\ast\}$, which is an arbitrary singleton set containing a single element $\ast$.

@@

            \subsubsection{Binary product}
            \begin{definition}[Binary product]
                Given a category $\mathcal{C}$, the \emph{binary product} of two objects $X$ and $Y$ in $\mathcal{C}$ is specified by
                \begin{itemize}
                    \item a $\mathcal{C}$-object $X \times Y$;
                    \item two $\mathcal{C}$-morphisms $\pi_1 : X \times Y \to X$ and $\pi_2 : X \times Y \to Y$ called the \emph{projections} of $X \times Y$;
                \end{itemize}
                such that for all $Z \in \textbf{obj}(\mathcal{C})$ and morphisms $f : Z \to X$ and $g : Z \to Y$, there exists a unique morphism $u : Z \to X \times Y$ such that the following diagram commutes in $\mathcal{C}$:
                \begin{equation}
                    \vcenter{\hbox{\includegraphics[width=0.4\textwidth]{binary_product.pdf}}}
                    \label{cd: binary_product}
                \end{equation}
                The unique morphism $u$ is written as $\ang{ f, g } : Z \to X \times Y$ where $f = \pi_1 \circ u$ and $g = \pi_2 \circ u$.
            \end{definition}
            It can be shown that the binary product is unique up to (unique) isomorphism.

@@

        \subsection{Functor}
        \begin{definition}[Functor]
            Given two categories $\mathcal{C}$ and $\mathcal{D}$, a \emph{functor} $F: \mathcal{C} \to \mathcal{D}$ is specified by
            \begin{itemize}
                \item 
                    a function 
                    \[\begin{aligned}
                        \textbf{obj}(\mathcal{C}) &\to \textbf{obj}(\mathcal{D}) \\
                        X &\mapsto F(X)
                    \end{aligned}\]

                \item 
                    for each $X, Y \in \textbf{obj}(\mathcal{C})$, a function 
                    \[\begin{aligned}
                        \mathcal{C}{(X,Y)} &\to \mathcal{D}{(F(X),F(Y))} \\
                        f &\mapsto F(f)
                    \end{aligned}\]
            \end{itemize}
            satisfying the following properties:
            \begin{itemize}
                \item 
                    For all $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, we have: $ F(\textbf{id}_X) = \textbf{id}_{F(X)}$
                \item
                    For all $X, Y, Z \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $g \in \mathcal{C}{(Y,Z)}$, we have: $F(g \circ f) = F(g) \circ F(f)$
            \end{itemize}
        \end{definition}

@@

        \subsection{Functor category}
        \begin{definition}[Functor category]
            Given two categories $\mathcal{C}$ and $\mathcal{D}$, the \emph{functor category} $\mathcal{D}^{\mathcal{C}}$ is the category satisfying the following:
            \begin{itemize}
                \item 
                    The objects of $\mathcal{D}^{\mathcal{C}}$ are all functors $\mathcal{C} \to \mathcal{D}$;
                \item 
                    Given two functors $F,G: \mathcal{C} \to \mathcal{D}$, the morphisms from $F$ to $G$ in $\mathcal{D}^{\mathcal{C}}$ are all natural transformations $\theta : F \to G$;
                \item
                    Composition and identity morphisms in $\mathcal{D}^{\mathcal{C}}$ are defined as follows:
                    \begin{itemize}
                        \item 
                            The identity morphism $\textbf{id}_F$ on $F$ is defined as $\theta_X = \textbf{id}_{F(X)}$ for all $X \in \textbf{obj}(\mathcal{C})$;
                        \item
                            The composition of two natural transformations $\theta : F \to G$ and $\phi : G \to H$ is defined as $(\phi \circ \theta)_X = \phi_X \circ \theta_X$ for all $X \in \textbf{obj}(\mathcal{C})$.
                    \end{itemize}
            \end{itemize}
        \end{definition}

        \subsection{Presheaf category}
        \begin{definition}[Presheaf]
            Given a category $\mathcal{C}$, a \emph{presheaf} on $\mathcal{C}$ is a functor $F: \mathcal{C}^{\textbf{op}} \to \mathcal{Set}$.
            A presheaf is a contravariant functor, which means that it reverses the direction of morphisms.
            In other words, a presheaf is a functor that takes objects in $\mathcal{C}$ and assigns them sets, and takes morphisms in $\mathcal{C}$ and assigns them functions between the corresponding sets.
            The presheaf $F$ is defined as follows:
            \begin{itemize}
                \item 
                    For each $X \in \textbf{obj}(\mathcal{C})$, $F(X)$ is a set;
                \item
                    For each $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $F(f)$ is a function $F(Y) \to F(X)$.
            \end{itemize}
        \end{definition}

        \begin{definition}[Presheaf category]
            Given a category $\mathcal{C}$, the \emph{presheaf category} $\hat{\mathcal{C}}$ is the functor category $\mathcal{Set}^{\mathcal{C}^{\textbf{op}}}$, which explicitly contains the following:
            \begin{itemize}
                \item 
                    The objects of $\hat{\mathcal{C}}$ are all presheaves on $\mathcal{C}$;
                \item
                    Given two presheaves $F,G: \mathcal{C}^{\textbf{op}} \to \mathcal{Set}$, the morphisms from $F$ to $G$ in $\hat{\mathcal{C}}$ are all natural transformations $\theta : F \to G$.
            \end{itemize}
        \end{definition}

        \subsection{Yoneda lemma}
        \begin{definition}[Yoneda functor]
            Given a category $\mathcal{C}$, the \emph{Yoneda functor} $\yo: \mathcal{C} \to \hat{\mathcal{C}}$ is defined as follows:
            \begin{itemize}
                \item 
                    For each $X \in \textbf{obj}(\mathcal{C})$, $\yo(X)$ is the functor $\mathcal{C}^{\textbf{op}} \to \mathcal{Set}$ defined as:
                    \begin{equation} \label{eq: yoneda_element}
                        \yo(X)(Y) = \mathcal{C}{(Y,X)}
                    \end{equation}
                    For all $Y \in \textbf{obj}(\mathcal{C})$;
                \item
                    For each $X, Y \in \textbf{obj}(\mathcal{C})$ and $f \in \mathcal{C}{(X,Y)}$, $\yo(f)$ is the morphism $\yo(X) \to \yo(Y)$ defined a natural transformation whose component at any given $Z \in \mathcal{C}^{\textbf{op}}$ is given by:
                    \begin{equation} \label{eq: yoneda_morphism}
                        \begin{aligned}
                            (\yo(f))_Z : \mathcal{C}{(Z,Y)} &\to \mathcal{C}{(Z,X)} \\
                            g &\mapsto g \circ f
                        \end{aligned}
                    \end{equation}
                    for all $Z \in \textbf{obj}(\mathcal{C})$.
                \end{itemize}

@@

        Here we define a datatype $\mb{\mathsf{Vec}}$ for vectors. The type of vectors is dependent on the length of the vector. With dependent types, we can encode properties directly into types.
